A Guide to Designing Experiments to Test Statistical Graphics

### 2022-07-05 index.qmd

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. 
Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.

>>>>

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. The missing boundary ellipse in the corresponding sub-panels escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.
